Loop-Aware Optimizations in PyPy's Tracing JIT

### 2011-06-29 talk/iwtc11/paper.tex
cleanups

\begin{abstract}
By introducing loop peeling into the optimization step of a tracing
jit the effect of optimizations already in place will be increased
greatly. Not only will they become able to move loop invariant code
out of loop. They will also become able to reuse results from the
previous iteration. Also, the implementation of excising optimizations
can be left almost intact as they will not have to interact much with
the loop peeling.

@@

Before the trace is passed to a backend compiling it into machine code
it needs to be optimized to achieve better performance.
The focus of this paper
is loop invariant code motion. The goal of that is to move as many
operations as possible out of the loop making them executed at most once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest some care has
to be taken when they are combined with loop peeling. This is
described below by first explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with the jump the the second part
(called the \emph{peeled loop}). The second part end with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.
Note that the peeled loop is not necessary the \emph{first} iteration of the
loop execution, it is general enough to correspond to any iteration of the loop.
However, the peeled loop can then be optimized using the assumption that a
previous iteration has happened.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the preamble, $J$, and jump
arguments $K$. Looking  at the peeled version of our example in Figure~\ref{fig:peeled-trace} we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second copy of the trace (the peeled loop) from the
first (the preeamble) we need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
copying. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

Each operation in the trace is copied in order.
To copy an operation $v=\text{op}\left(A_1, A_2, \cdots, A_{|A|}\right)$
a new variable, $\hat v$ is introduced. The copied operation will
return $\hat v$ using
\begin{equation}
  \hat v = \text{op}\left(m\left(A_1\right), m\left(A_2\right), 
    \cdots, m\left(A_{|A|}\right)\right) . 
\end{equation}
Before the
next operation is copied, $m$ is extend by assigning $m\left(v\right) = \hat
v$. For the example above, after all the operations have been copied we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5 \\
      m\left(i_2\right) &=& i_6 \\
      m\left(i_3\right) &=& i_7 \\
      m\left(i_4\right) &=& i_8 \\
      m\left(p_5\right) &=& p_9 \\
    \end{array}
  %\right.
  .
\end{equation}

The trace from Figure~\ref{fig:unopt-trace} would after this operation become
the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
preamble while line 15-27 shows the peeled loop.

@@

In general, after loop peeling and redundant operation removal the peeled loop
will no longer be in SSA form as it operates on variables that are the result
of pure operations in the preamble. The solution is to extend the input
arguments, $J$, with those variables. This will also extend the
jump arguments of the preamble, which is also $J$. 
Implicitly that also extends the jump arguments of the peeled loop, $K$,
since they are the image of $J$ under $m$. For the example $I$ has to
be replaced by $\hat I$ which is formed by appending $i_3$ to $I$.
At the same time $K$ has to be replaced by
$\hat K$ which is formed by appending $m\left(i_3\right)=i_7$ to $K$.
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$.

In general what is needed is to keep track of
which variables from the preamble it reuses in the peeled loop.
It has to construct a vector, $H$,  of such variables which
can be used to update the input and jump arguments using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$.

@@

PyPy's allocation removal optimization \cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. Those objects have to be allocated in the loop, but no outside
object ever gets a reference short lived objects with no references outside the
loop. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

When the optimizer reaches line 13 it needs to construct the
arguments of the \lstinline{jump} operation, which contains the
reference to the allocation-removed object in $p_5$. This can be achieved by
exploding $p_5$ into the fields of the allocation-removed object.
In this case there is only one such field and its value is
$i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
fields\footnote{This is sometimes called \emph{scalar replacement}.}.
If some of the fields are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the fields in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
would be short numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

The effect of combining a one ass optimization with loop peeling gives
completely standard loop invariant code motion optimizations
\cite{muchnick_advanced_1997}. We do not claim any novelty in the effect, but
think that our implementation scheme is a very simple one.

@@

Both the Hotpath VM \cite{gal_hotpathvm:_2006} and SPUR
\cite{bebenita_spur:_2010} implements loop-invariant code motion
directly, by explicitly marking as loop-invariant all variables that stay the
same along all looping paths and then moving all pure computation that depends
only on these variables out of the loop. SPUR can also hoist loads out of the
loop if nothing in the loop can ever write to the memory location. It can also
move allocations out of the loop, but does not replace the object by its fields.
This saves only the allocation, not the access to the object fields.

>>>>

\begin{abstract}
By introducing loop peeling into the optimization step of a tracing
jit the effect of optimizations already in place will be increased
greatly. Not only does it make them move loop invariant code
out of loop. They will also become able to reuse results from the
previous iteration. Also, the implementation of excising optimizations
can be left almost intact as they will not have to interact much with
the loop peeling.

@@

Before the trace is passed to a backend compiling it into machine code
it is optimized to achieve better performance.
One goal of that is to move 
operations out of the loop making them executed only once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest some care has
to be taken when they are combined with loop peeling. This is
described below by explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

Loop peeling is achieved by appending an copy of the traced iteration at
the end of itself. See Figure~\ref{fig:overview} for an illustration.
The first part (called \emph{preamble}) finishes with a jump the the second part
(called the \emph{peeled loop}). The second part finishes with a jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every further iteration. New variable names have to be
introduced in the entire copied trace in order to maintian the SSA-property.
Note that the peeled loop is not necessary the \emph{first} iteration of the
loop execution, it is general enough to correspond to any iteration of the loop.
However, the peeled loop can then be optimized using the assumption that a
previous iteration has happened.

@@

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the preamble, $J$, and jump
arguments $K$. 
Figure~\ref{fig:overview} illustrates the general case. The running
example in Figure~\ref{fig:unopt-trace} has  $I = \left( p_0, p_1
\right)$ and $J = \left( p_0, p_5 \right)$. The result of applying
loop peeling to it is shown in Figure~\ref{fig:peeled-trace} with 
$K = \left( p_0, p_9 \right)$.

To construct the second copy of the trace (the peeled loop) from the
first (the preeamble) we need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
copying. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

Each operation in the trace is copied in order.
To copy an operation $v=\text{op}\left(A_1, A_2, \cdots, A_{|A|}\right)$
a new variable, $\hat v$ is introduced. The copied operation will
return $\hat v$ using
\begin{equation}
  \hat v = \text{op}\left(m\left(A_1\right), m\left(A_2\right), 
    \cdots, m\left(A_{|A|}\right)\right) . 
\end{equation}
Before the
next operation is copied, $m$ is extend by assigning $m\left(v\right) = \hat
v$. For the example above, that will extend $m$ with
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(i_2\right) &=& i_6 \\
      m\left(i_3\right) &=& i_7 \\
      m\left(i_4\right) &=& i_8 \\
      m\left(p_5\right) &=& p_9 \\
    \end{array}
  %\right.
  .
\end{equation}

@@

After loop peeling and redundant operation removal the peeled loop
will typically no longer be in SSA form but operate on variables that are the result
of operations in the preamble. The solution is to extend the input
arguments, $J$, with those variables. This will also extend the
jump arguments of the preamble, which is also $J$. 
Implicitly that also extends the jump arguments of the peeled loop, $K$,
since they are the image of $J$ under $m$. For the example $I$ has to
be replaced by $\hat I$ which is formed by appending $i_3$ to $I$.
At the same time $K$ has to be replaced by
$\hat K$ which is formed by appending $m\left(i_3\right)=i_7$ to $K$.
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$.

In general what is needed is to keep track of
which variables from the preamble are reused in the peeled loop.
By constructing a vector, $H$,  of such variables, the input and jump
arguments can be updated using
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
  \label{eq:heap-inputargs}
\end{equation}
and
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
  \label{eq:heap-jumpargs}
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$.

@@

PyPy's allocation removal optimization \cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. That is, no outside
object ever gets a reference to them. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

When the optimizer reaches line 13 it needs to construct the
arguments of the \lstinline{jump} operation, which contains the
reference to the allocation-removed object in $p_5$. This can be achieved by
exploding $p_5$ into the attributes of the allocation-removed object.
In this case there is only one such attribute and its value is
$i_4$, which means that $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes\footnote{This is sometimes called \emph{scalar replacement}.}.
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
is short and contains numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes runtime performance worse, in
the worst case the peeled loop is exactly the same as the preamble. Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

The effect of combining a one pass optimization with loop peeling gives
completely standard loop invariant code motion optimizations
\cite{muchnick_advanced_1997}. We do not claim any novelty in the effect, but
think that our implementation scheme is a very simple one.

@@

Both the Hotpath VM \cite{gal_hotpathvm:_2006} and SPUR
\cite{bebenita_spur:_2010} implements loop-invariant code motion
directly, by explicitly marking as loop-invariant all variables that stay the
same along all looping paths and then moving all pure computation that depends
only on these variables out of the loop. SPUR can also hoist loads out of the
loop if nothing in the loop can ever write to the memory location. It can also
move allocations out of the loop, but does not replace the object by its attributes.
This saves only the allocation, not the access to the object attributes.
